# Women's Basketball Officials Analysis
### Derek Willis
### IRE 2025 - Storyball: Finding Ideas in Sports Data

Building off the website showing the schedules of NCAA men's college basketball officials, let's take a look at women's basketball officials from the 2024-25 season, focusing on how many days the most active ones worked. The following data has a row for each day an official worked.

```{r}
library(tidyverse)
library(lubridate)

officials <- read_csv("officials_202425.csv") |> mutate(month = month(date), day = day(date))

glimpse(officials)

```
### Exploration

Let's start by finding the officials who worked the most games. Bear in mind that this doesn't include some teams' games, so it's not comprehensive, but it does represent most of the NCAA contests last season. How does the result compare with the [top men's officials](https://blessyourchart.shinyapps.io/cbb-ref-logs/)? What about the total number of women's referees compared to the men's?

```{r}
officials |> 
  group_by(official) |> 
  summarize(games = n()) |> 
  arrange(desc(games))
```

Let's do some descriptive statistics:

```{r}
ref_counts <- officials |> 
  group_by(official) |> 
  summarize(games = n())

ref_counts |> 
  summarize(
    mean = mean(games),
    median = median(games)
  )

```
Seems like quite a few officials are only calling a small number of games, meaning that a handful of refs are calling a lot of games.

```{r}
ref_counts |> 
  ggplot() +
  geom_histogram(aes(x=games))
```

What about the officials with the most active months?

```{r}
officials |> 
  group_by(official, month) |> 
  summarize(games = n()) |> 
  arrange(desc(games))
```

John Capolino had a pretty busy November and December, working a combined 41 days in that 62-day span (which often includes holiday tournaments in both months). What about streaks of consecutive days working? I asked my buddy Claude to write R code "that finds the official who worked the longest stretch of consecutive days, and how many times they did that."

```{r}
# Function to find consecutive date streaks
find_consecutive_streaks <- function(dates) {
  if(length(dates) == 0) return(integer(0))
  
  # Sort dates
  sorted_dates <- sort(dates)
  
  # Find breaks in consecutive dates
  date_diffs <- c(1, diff(sorted_dates))
  break_points <- which(date_diffs > 1)
  
  # Calculate streak lengths
  if(length(break_points) == 0) {
    # All dates are consecutive
    return(length(sorted_dates))
  } else {
    # Multiple streaks
    streak_starts <- c(1, break_points)
    streak_ends <- c(break_points - 1, length(sorted_dates))
    streak_lengths <- streak_ends - streak_starts + 1
    return(streak_lengths)
  }
}

# Function to analyze consecutive days for all officials
analyze_consecutive_days <- function(data) {
  # Convert date column to Date type
  data$date <- as.Date(data$date)
  
  # Get unique dates for each official
  official_dates <- data |>
    group_by(official) |>
    summarise(dates = list(unique(date)), .groups = 'drop')
  
  # Calculate streaks for each official
  results <- official_dates |>
    rowwise() |>
    mutate(
      all_streaks = list(find_consecutive_streaks(dates)),
      max_streak = ifelse(length(all_streaks) > 0, max(all_streaks), 0),
      total_days = length(dates),
      num_streaks = length(all_streaks),
      times_achieved_max = sum(all_streaks == max_streak)
    ) |>
    select(-dates, -all_streaks) |>
    ungroup()
  
  # Find the overall longest streak
  overall_max <- max(results$max_streak)
  
  # Add column indicating if this official has the longest streak
  results <- results |>
    mutate(
      has_longest_streak = max_streak == overall_max,
      overall_longest_streak = overall_max
    ) |>
    arrange(desc(max_streak), desc(times_achieved_max), official)
  
  return(results)
}

# Analyze the data
results_df <- analyze_consecutive_days(officials)

# Display results
print("=== Consecutive Days Analysis Results ===")
print(results_df)

# Summary of officials with longest streak
longest_streak_officials <- results_df |>
  filter(has_longest_streak == TRUE) |>
  select(official, max_streak, times_achieved_max)

cat("\n=== Officials with Longest Consecutive Streak ===\n")
cat("Longest streak:", max(results_df$max_streak), "days\n\n")

for(i in 1:nrow(longest_streak_officials)) {
  cat(sprintf("%s: %d consecutive days (achieved %d time%s)\n", 
              longest_streak_officials$official[i],
              longest_streak_officials$max_streak[i],
              longest_streak_officials$times_achieved_max[i],
              ifelse(longest_streak_officials$times_achieved_max[i] == 1, "", "s")))
}

# Optional: Create a more detailed dataframe with individual streak information
create_detailed_streaks <- function(data) {
  data$date <- as.Date(data$date)
  
  detailed_results <- data |>
    group_by(official) |>
    summarise(dates = list(unique(sort(date))), .groups = 'drop') |>
    rowwise() |>
    mutate(
      streaks = list(find_consecutive_streaks(dates)),
      streak_details = list(data.frame(
        streak_number = seq_along(streaks),
        streak_length = streaks
      ))
    ) |>
    select(official, streak_details) |>
    tidyr::unnest(streak_details)
  
  return(detailed_results)
}

# Create detailed streaks dataframe
detailed_df <- create_detailed_streaks(sample_data)

cat("\n=== Detailed Streaks by Official ===\n")
print(detailed_df)

# Return the main results dataframe
results_df
```

Teresa Stuck, you need a break! Let's see when that 13-day stretch happened.

```{r}
officials |> 
  filter(official == 'Teresa Stuck') |> 
  arrange(date)
```

So from Nov. 11 through Nov. 23, she worked a game every day, going from Boston to Stony Brook to Washington to Starkville, MS on consecutive days. Here's the potential investigative angle: do those kinds of stretches impact officiating? How could we tell? 
