# Women's Basketball Officiating Analysis
### Derek Willis
### IRE 2025 - Storyball: Finding Ideas in Sports Data

## Setup

Load the required packages and data:

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)

# Load the data
wbb_data <- read_csv("wbblogs.csv")

# Quick look at the data structure
glimpse(wbb_data)
```

## Understanding Fouls in the Data

We have two foul metrics:
- `TeamPersonalFouls`: fouls called against the team
- `OpponentPersonalFouls`: fouls called against opponents

```{r data_check}
# Check for missing values in foul columns
wbb_data |> 
  summarize(
    team_fouls_missing = sum(is.na(TeamPersonalFouls)),
    opponent_fouls_missing = sum(is.na(OpponentPersonalFouls)),
    total_games = n()
  )
```

## Fouls by Conference

Which conferences have different officiating patterns?

```{r conference_fouls}
conference_fouls <- wbb_data |> 
  group_by(Conference) |> 
  summarize(
    games = n(),
    avg_team_fouls = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_opponent_fouls = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  mutate(
    foul_differential = avg_team_fouls - avg_opponent_fouls,
    total_avg_fouls = avg_team_fouls + avg_opponent_fouls
  ) |> 
  arrange(desc(total_avg_fouls))

# Show top 15 conferences by total fouls per game
conference_fouls |> 
  head(15) |> 
  kable(caption = "Conferences with Most Fouls Called Per Game")
```

#### What should we make of these results? What do the top two conferences have in common?


```{r conference_discipline}
# Which conferences have the biggest foul differentials?
conference_fouls |> 
  arrange(desc(abs(foul_differential))) |> 
  head(15) |> 
  select(Conference, games, avg_team_fouls, avg_opponent_fouls, foul_differential) |> 
  kable(caption = "Biggest Officiating Imbalances by Conference")
```

```{r conference_low_fouls}
# Conferences with fewest fouls called
conference_fouls |> 
  arrange(total_avg_fouls) |> 
  head(10) |> 
  select(Conference, games, total_avg_fouls, avg_team_fouls, avg_opponent_fouls) |> 
  kable(caption = "Conferences with Fewest Fouls Called")
```

## Team-Level Officiating Patterns

Find teams that consistently see different foul calling:

```{r team_fouls}
team_fouls <- wbb_data |> 
  group_by(Team, Conference) |> 
  summarize(
    games = n(),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    conference = first(Conference),
    .groups = "drop"
  ) |> 
  filter(games >= 10) |>  # Only teams with substantial game counts
  mutate(
    foul_differential = avg_fouls_against_team - avg_fouls_against_opponent,
    total_fouls = avg_fouls_against_team + avg_fouls_against_opponent
  )

# Teams with most fouls called against them
team_fouls |> 
  arrange(desc(avg_fouls_against_team)) |> 
  head(15) |> 
  select(Team, Conference, games, avg_fouls_against_team) |> 
  kable(caption = "Teams With Most Fouls Called Against Them")
```

```{r fewest_fouls}
# Teams with fewest fouls called against them
team_fouls |> 
  arrange(avg_fouls_against_team) |> 
  head(15) |> 
  select(Team, Conference, games, avg_fouls_against_team) |> 
  kable(caption = "Teams With Fewest Fouls Called Against Them")
```

```{r officiating_advantage}
# Teams that get the most favorable officiating (big negative differential)
team_fouls |> 
  arrange(foul_differential) |> 
  head(15) |> 
  select(Team, Conference, games, avg_fouls_against_team, avg_fouls_against_opponent, foul_differential) |> 
  kable(caption = "Teams With Most Favorable Foul Differential")
```

```{r officiating_disadvantage}
# Teams that get the least favorable officiating (big positive differential)
team_fouls |> 
  arrange(desc(foul_differential)) |> 
  head(15) |> 
  select(Team, Conference, games, avg_fouls_against_team, avg_fouls_against_opponent, foul_differential) |> 
  kable(caption = "Teams With Least Favorable Foul Differential")
```

## HBCU Officiating Analysis

Do officials call games differently for HBCUs?

```{r hbcu_fouls}
hbcu_fouls <- wbb_data |> 
  group_by(hbcu) |> 
  summarize(
    games = n(),
    teams = n_distinct(Team),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  mutate(
    foul_differential = avg_fouls_against_team - avg_fouls_against_opponent,
    total_fouls_per_game = avg_fouls_against_team + avg_fouls_against_opponent
  )

hbcu_fouls |> 
  kable(caption = "Officiating Patterns: HBCU vs Non-HBCU")
```

**Key question**: Are HBCUs getting different treatment from officials?

```{r hbcu_detailed}
# Break down by individual HBCU teams
hbcu_teams <- wbb_data |> 
  filter(hbcu == TRUE) |> 
  group_by(Team, Conference) |> 
  summarize(
    games = n(),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  filter(games >= 10) |> 
  mutate(foul_differential = avg_fouls_against_team - avg_fouls_against_opponent) |> 
  arrange(desc(abs(foul_differential)))

hbcu_teams |> 
  head(20) |> 
  kable(caption = "HBCU Teams with Largest Officiating Differentials")
```

```{r hbcu_vs_conference}
# Compare HBCUs to their conference averages
hbcu_conference_comparison <- wbb_data |> 
  filter(hbcu == TRUE) |> 
  group_by(Team, Conference) |> 
  summarize(
    games = n(),
    team_avg_fouls = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    team_opp_fouls = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  filter(games >= 10) |> 
  left_join(conference_fouls, by = "Conference", suffix = c("", "_conf")) |> 
  mutate(
    fouls_vs_conf_avg = team_avg_fouls - avg_team_fouls,
    opp_fouls_vs_conf_avg = team_opp_fouls - avg_opponent_fouls
  ) |> 
  select(Team, Conference, games, team_avg_fouls, avg_team_fouls, fouls_vs_conf_avg, 
         team_opp_fouls, avg_opponent_fouls, opp_fouls_vs_conf_avg)

hbcu_conference_comparison |> 
  arrange(desc(abs(fouls_vs_conf_avg))) |> 
  head(15) |> 
  kable(caption = "How HBCU Teams Compare to Conference Averages")
```

## Home vs Away Officiating

Do officials favor home teams in foul calling?

```{r home_away}
home_away_fouls <- wbb_data |> 
  group_by(HomeAway) |> 
  summarize(
    games = n(),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  mutate(foul_differential = avg_fouls_against_team - avg_fouls_against_opponent)

home_away_fouls |> 
  kable(caption = "Home Court Advantage in Officiating")
```

```{r home_away_by_conference}
# Home court advantage by conference
home_court_by_conf <- wbb_data |> 
  group_by(Conference, HomeAway) |> 
  summarize(
    games = n(),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  mutate(foul_differential = avg_fouls_against_team - avg_fouls_against_opponent) |> 
  pivot_wider(names_from = HomeAway, values_from = c(games, avg_fouls_against_team, 
                                                     avg_fouls_against_opponent, foul_differential)) |> 
  mutate(
    home_advantage = foul_differential_Away - foul_differential_Home
  ) |> 
  arrange(desc(home_advantage))

home_court_by_conf |> 
  head(15) |> 
  select(Conference, home_advantage, foul_differential_Home, foul_differential_Away) |> 
  kable(caption = "Biggest Home Court Advantages in Officiating by Conference")
```

## Win/Loss Impact

Does officiating correlate with game outcomes?

```{r win_loss_fouls}
win_loss_fouls <- wbb_data |> 
  group_by(W_L) |> 
  summarize(
    games = n(),
    avg_fouls_against_team = round(mean(TeamPersonalFouls, na.rm = TRUE), 1),
    avg_fouls_against_opponent = round(mean(OpponentPersonalFouls, na.rm = TRUE), 1),
    .groups = "drop"
  ) |> 
  mutate(foul_differential = avg_fouls_against_team - avg_fouls_against_opponent)

win_loss_fouls |> 
  kable(caption = "Foul Calling Patterns in Wins vs Losses")
```

```{r extreme_foul_games}
# Games with extreme foul differentials
extreme_fouls <- wbb_data |> 
  mutate(
    foul_diff = TeamPersonalFouls - OpponentPersonalFouls,
    abs_foul_diff = abs(foul_diff)
  ) |> 
  filter(abs_foul_diff >= 15) |> 
  arrange(desc(abs_foul_diff)) |> 
  select(Date, Team, Opponent, HomeAway, W_L, TeamPersonalFouls, OpponentPersonalFouls, 
         foul_diff, Conference, hbcu)

extreme_fouls |> 
  head(20) |> 
  kable(caption = "Games with Extreme Foul Differentials (15+ difference)")
```

## Season-Long Patterns

Are there trends over time?

```{r season_patterns}
# Foul patterns by season
season_fouls <- wbb_data |> 
  group_by(Season) |> 
  summarize(
    games = n(),
    avg_total_fouls = round(mean(TeamPersonalFouls + OpponentPersonalFouls, na.rm = TRUE), 1),
    avg_foul_differential = round(mean(abs(TeamPersonalFouls - OpponentPersonalFouls), na.rm = TRUE), 1),
    games_with_big_diff = sum(abs(TeamPersonalFouls - OpponentPersonalFouls) >= 10, na.rm = TRUE),
    pct_lopsided = round(100 * games_with_big_diff / games, 1),
    .groups = "drop"
  ) |> 
  arrange(Season)

season_fouls |> 
  kable(caption = "Officiating Trends by Season")
```

## Statistical Significance Tests

```{r statistical_tests}
# Test if HBCU foul differential is significantly different
hbcu_test <- wbb_data |> 
  mutate(
    foul_diff = TeamPersonalFouls - OpponentPersonalFouls,
    is_hbcu = ifelse(is.na(hbcu), "No", "Yes")  # Convert TRUE/NA to Yes/No
  ) |> 
  filter(!is.na(foul_diff))

# Check the levels first
cat("HBCU levels in data:", unique(hbcu_test$is_hbcu), "\n")
cat("Count by HBCU status:\n")
print(table(hbcu_test$is_hbcu))

t_test_result <- t.test(foul_diff ~ is_hbcu, data = hbcu_test)

cat("\nHBCU vs Non-HBCU Foul Differential T-Test:\n")
cat("HBCU mean foul differential:", round(t_test_result$estimate[2], 3), "\n")
cat("Non-HBCU mean foul differential:", round(t_test_result$estimate[1], 3), "\n")
cat("Difference (HBCU - Non-HBCU):", round(t_test_result$estimate[2] - t_test_result$estimate[1], 3), "\n")
cat("P-value:", round(t_test_result$p.value, 6), "\n")
cat("95% Confidence Interval:", round(t_test_result$conf.int, 3), "\n")

# Test home court advantage
home_test <- wbb_data |> 
  mutate(foul_diff = TeamPersonalFouls - OpponentPersonalFouls) |> 
  filter(!is.na(foul_diff), HomeAway != "Neutral")

home_t_test <- t.test(foul_diff ~ HomeAway, data = home_test)

cat("\nHome vs Away Foul Differential T-Test:\n")
cat("Mean difference:", round(home_t_test$estimate[2] - home_t_test$estimate[1], 3), "\n")
cat("P-value:", round(home_t_test$p.value, 6), "\n")
```

The results suggest there's a systematic difference in how fouls are distributed in HBCU versus non-HBCU games, with HBCU games showing more uneven foul calling. The question is why.
